% !TeX spellcheck = en_US
\section{Decision Problems}

\subsection{Fundamental definitions}

A decision problem is defined by a 6-uple: 
$$ P = (X, \Omega, F, f, D, \Pi)$$
Where
\begin{itemize}
	\item $X$ is the feasible region, set of all alternatives
	
	\item $\Omega$ is the sample space, set of all possible scenarios
	
	\item $F$ is the indicator space, set of all possible impacts 
	
	\item $f: X \times \Omega \rightarrow F$ is the impact function, which associates each possible configuration $f(x, \omega)$ of the system to an impact
	
	\item $D$ is the set of all decision-makers
	
	\item $\Pi: D \rightarrow 2^{F \times F}$ is the preference function, which associates to each decision maker a subset of impact pairs
\end{itemize}

The \textbf{alternatives} formally describe the events under the control of the decision-makers. The feasible region $X$ includes all possible choices. 

Alternatives are described as vectors of real numbers, each component of the alternative is a \textbf{decision variable}, an elementary quantity describing a specific aspect of the alternative. The core of the decision process is choosing an alternative with the best possible outcome (impact), the decision variables create the decision space (all the values that an alternative can assume), and solutions are point within that space that have to be evaluated by the decision process.

The \textbf{scenarios} formally describe the events out of the control of the decision-makers. The sample space $\Omega$ defines all such events.

The \textbf{impacts} model all aspects relevant to the decision, they are described quantitatively as vector of real numbers.

The \textbf{impact function} is a vectorial function which associates each configuration to an impact.

A \textbf{decision-maker} is whoever takes part in the decision. 

The decision-maker takes part directly in the choice of alternative, holding the power and responsibility to decide, while \textbf{stakeholders} are all the subjects who don't actively participate in the decision but whose interests are affected by its outcome.

\subsection{Preference Relation}

For each decision maker, we need a relation between pair of impacts to determine which ones are preferable, $\Pi: D \rightarrow 2^{F \times F}$ associates each decision-maker to a subset of impact pairs.

A \textbf{weak preference} $f \wpref{d} f'$ is when the decision-maker accepts the exchange of $f$ for $f'$.

Two impacts are \textbf{indifferent} if the decision-maker considers both equally as satisfactory, i.e., is willing to exchange one for the other in both directions.

Two impacts are \textbf{incomparable} if the decision-maker is unable or unwilling to choose between them, rejecting the exchange in both directions.

In a decision problem, an \textbf{outperformer} $x \in X$ for $d \in D$ is an alternative which performs at least as good as any other $x' \in X$ in every scenario $\omega \in \Omega$
$$ (f(x, \omega), f(x', \omega)) \in \Pi_d, \quad \forall \omega \in \Omega, \ \forall x' \in X $$
and is strictly better for at least one comparison/scenario
$$ \exists x^\circ \in X \quad \left(f(x, \omega), f(x^\circ, \omega)\right) \in \Pi_d \wedge \left(f(x^\circ, \omega), f(x, \omega)\right) \notin \Pi$$

\subsubsection{Properties}

Considering impact set $F$ and a binary relation $\Pi$, the main properties of said relation can be:
\begin{itemize}
	\item \textbf{Reflexivity:} $\forall a \in F$, $(a,a) \in \Pi$, every impact is in relation with itself; to check for it, on the graph representation every impact has to have a self-loop
	
	\item \textbf{Transitivity:} $\forall a,b,c \in F$, if $(a,b) \in \Pi$ and $(b,c) \in \Pi$ then $(a,c) \in \Pi$; to check for it, on the graph representation follow every arc and check that every possible "triangle" is complete
	
	\item \textbf{Antisymmetry:} $\forall a,b \in F$, if $(a,b) \in \Pi$ and $(b,a) \in \Pi$ then $a = b$, two impacts are indifferent only if they're equal; to check for it, on the graph representation there must be no impacts that point at each other
	
	\item \textbf{Completeness:} $\forall a,b \in F$, if $(a,b) \notin \Pi$ then $(b, a) \in \Pi$, every impact is in relation with every other, one way or another; to check for it, on the graph representation, from each node there must be an arc (either outgoing or incoming) connecting it to every other
\end{itemize}

\subsubsection{Types of relations (orders)}

Combining the properties, we can get different kinds of preferences: 
\begin{itemize}
	\item \textbf{Preorder:} reflexivity and transitivity. Guarantees that the set of nondominated alternatives is nonempty, even if not all alternatives are comparable, we can find at least one nondominated solution
	
	\item \textbf{Partial order:} reflexivity, transitivity and antisymmetry. This limits indifference, but still allows for incomparability, and thus not always leading to a definitive choice
	
	\item \textbf{Weak order:} reflexivity, transitivity and completeness. Guarantees that nondominated solutions exists and are all mutually indifferent, any of them can be chosen. Such orders admit representation by a value function (with ties), turning the decision problem into an optimization problem
	
	\item \textbf{Total order:} reflexivity, transitivity, antisymmetry and completeness. Provides a unique linear ranking of alternatives, there is always a unique best alternative
\end{itemize}

\subsubsection{Derived relations}

From the weak preference relation, one can derive: 
\begin{itemize}
	\item \textbf{Indifference relation:} $\indiff{\Pi}$
	$$ (a,b),(b,a) \in \indiff{\Pi} \Leftrightarrow (a,b) \in \Pi \wedge (b,a) \in \Pi $$
	To build it from the graph representation, add all self loops and each pair of arcs that point at each other
	
	\item \textbf{Strict preference relation:} $\spref{\Pi}$
	$$ (a,b) \in \spref{\Pi} \Leftrightarrow (a,b) \in \Pi \wedge (b,a) \notin \Pi $$
	To build it from the graph representation, add all arcs which do not have an equal one in the opposite direction (the set difference of the last one w.r.t. $\Pi$)
	
	\item \textbf{Incomparability relation:} $\inc{\Pi}$
	$$ (a,b), (b,a) \in \inc{\Pi} \Leftrightarrow (a,b) \notin \Pi \wedge (b,a) \notin \Pi $$
	To build it from the graph representation, add all arcs which are not present in either direction in the graph
\end{itemize}